---
title: "xcms3_preprocessing"
author: "Daria Rapoport"
date: "13 01 2020"
output: html_document
---
## Install required packages

```{r, eval = FALSE, message = FALSE}
if (!requireNamespace("BiocManager"))
    install.packages("BiocManager")
BiocManager::install(c("xcms", "CAMERA"))
```

## Preprocess your data using XCMS3 and export data files for feature-based molecular networking through GNPS


Load required libraries and utility functions for GNPS export.

```{r, message = FALSE}
library(xcms)
source("https://raw.githubusercontent.com/jorainer/xcms-gnps-tools/master/customFunctions.R")
```

The number (`4`) defines the number of parallel tasks. Adapt this setting to the number of CPUs
available on your system (a bit less)

```{r}
if (.Platform$OS.type == "unix") {
  register(bpstart(MulticoreParam(6)))
} else {
  register(bpstart(SnowParam(6)))
} 
```

### Load data



```{r}


mzMLfiles <- paste0('/home/ldem/Documents/Daria/2021/LCMS/interactions/',
                    list.files(path = '/home/ldem/Documents/Daria/2021/LCMS/interactions',
                               pattern = ".mzML$", recursive = TRUE))

mzMLfiles=mzMLfiles[c(3,11)]



s_groups <- rep("sample", length(mzMLfiles))
pheno <- data.frame(sample_name = sub(basename(mzMLfiles), pattern = ".mzML",replacement = "", fixed = TRUE),
                    sample_group = s_groups, stringsAsFactors = FALSE)
```


Read all raw data (which includes MS1 and MS2 spectra).

```{r}
rawData <- readMSData(mzMLfiles, centroided. = TRUE, mode = "onDisk",
                      pdata = new("NAnnotatedDataFrame", pheno))
```

Create a base peak chromatogram (BPC) for visual inspection.

```{r, fig.width = 12, fig.height = 6, fig.cap = "Base peak chromatogram."}
bpis <- chromatogram(rawData, aggregationFun = "max")
plot(bpis)
head(rtime(rawData))


```

### Peak picking

Define settings for the centWave peak detection. As mentioned in the
introduction, these settings should always be adapted to the analyzed data set.

```{r}
cwp <- CentWaveParam(snthresh = 5, noise = 1000, peakwidth = c(3, 30), ppm = 20)

###if like in xcms online:
#cwp <- CentWaveParam(snthresh = 6, noise = 0, peakwidth = c(5, 20), ppm = 15)
```

Perform the chromatographic peak detection using *centWave*.

```{r, warning = FALSE, message = FALSE}
processedData <- findChromPeaks(rawData, param = cwp)
```

Get an overview of the detected peaks, using a heatmap which represents the
number of peaks detected for each file along the retention time range.

```{r, fig.width = 10, fig.height = 6}
plotChromPeakImage(processedData, binSize = 10) 
```

### Retention time alignment

Do an *obiwarp*-based alignment using the default settings (ideally adapt
settings to the analyzed data set).

```{r, message = FALSE, warning = FALSE}
processedData <- adjustRtime(processedData, param = ObiwarpParam())
```

Plot the difference between adjusted and raw retention times.

```{r, fig.width = 12, fig.height = 6}
plotAdjustedRtime(processedData)
```


### Peak grouping

Define the parameters for the *peak density*-based peak grouping (correspondence
analysis).

```{r, message = FALSE, warning = FALSE}
pdp <- PeakDensityParam(sampleGroups = processedData$sample_group,
                        minFraction = 0.50)
processedData <- groupChromPeaks(processedData, param = pdp) 
```

### Gap filling

Fill-in missing peaks. Peak detection might have failed for some features in
some samples. The `fillChromPeaks` function allows to integrate for such cases
all signal in the respective m/z - retention time range. Below we first define
the median width of identified chromatographic peaks in retention time dimension
and use this as parameter `fixedRt` for the `fillChromPeaks`.

```{r, message = FALSE, warning = FALSE}
medWidth <- median(chromPeaks(processedData)[, "rtmax"] -
                   chromPeaks(processedData)[, "rtmin"])


## fill missing peaks

###!!!!NB I had a mistake here!!!! processed_Data
processedData <- fillChromPeaks(
    processedData, param = FillChromPeaksParam(fixedRt = medWidth))
```

### Export data

#### export MS1 and MS2 features

Below we use the `featureSpectra` function to extract all MS2 spectra with their
precursor m/z being within the m/z range of a feature/peak and their retention
time within the rt range of the same feature/peak. Zero-intensity
values are removed from each spectrum with the `clean` function, and
subsequently processed into the expected format using the `formatSpectraForGNPS`
function.

```{r}
## export the individual spectra into a .mgf file

#memory.size()
#memory.limit()
# memory.limit(size=56000)

filteredMs2Spectra <- featureSpectra(processedData, return.type = "MSpectra")
filteredMs2Spectra <- clean(filteredMs2Spectra, all = TRUE)
filteredMs2Spectra <- formatSpectraForGNPS(filteredMs2Spectra)
```

The extracted MS2 spectra are saved as *ms2spectra_all.mgf* file. This file can
for example be used to do *in silico* structure prediction through
[SIRIUS+CSI:FingerID](https://bio.informatik.uni-jena.de/software/sirius/).

```{r}
writeMgfData(filteredMs2Spectra, "ms2spectra_all_corr0801.mgf")
```

Export peak area quantification table. To this end we first extract the *feature
definitions* (i.e. the m/z and retention time ranges and other metadata for all
defined features in the data set) and then the integrated peak areas (with the
`featureValues` function). This peak area quantification table contains features
and respective per sample peak areas in columns. The combined data is then saved
to the file *xcms_all.txt*. Note that it is now also possible to use the entire
feature table in the FBMN workflow.

```{r}
## get feature definitions and intensities
featuresDef <- featureDefinitions(processedData)
featuresIntensities <- featureValues(processedData, value = "into")

## generate data table
dataTable <- merge(featuresDef, featuresIntensities, by = 0, all = TRUE)
dataTable <- dataTable[, !(colnames(dataTable) %in% c("peakidx"))]
```

```{r}
head(dataTable)
```

```{r}
write.table(dataTable, "xcms_all_corr0801.txt", sep = "\t", quote = FALSE, 
            row.names = FALSE)
```


#### Export MS2 features only

The `filteredMs2Spectra` contains all MS2 spectra with their precursor m/z
within the feature's m/z range and a retention time that is within the retention
time of the chromatographic peak/feature. We thus have multiple MS2 spectra for
each feature (also from each sample). Metadata column `"feature_id"` indicates
to which feature a MS2 spectrum belongs:

```{r}
filteredMs2Spectra
```

We next select a single MS2 spectrum for each feature and export this reduced
set also as an .mgf file. We use the `combineSpectra` function on the list of
spectra and specify with `fcol = "feature_id"` how the spectra are grouped
(i.e. all spectra with the same feature id are processed together). On the set
of spectra of the same feature we apply the `maxTic` function that simply
returns the spectrum with the largest sum of intensities. We thus select with
the code below the spectrum with the largest total signal as the
*representative* MS2 spectrum for each feature.

```{r}
## Select for each feature the Spectrum2 with the largest TIC.
filteredMs2Spectra_maxTic <- combineSpectra(filteredMs2Spectra,
                                            fcol = "feature_id",
                                            method = maxTic)
```

Next we export the data to a file which can then be submitted to GNPS [feature-based
molecular
networking](https://ccms-ucsd.github.io/GNPSDocumentation/featurebasedmolecularnetworking/).

```{r}
writeMgfData(filteredMs2Spectra_maxTic, "ms2spectra_maxTic_corr0801.mgf")
```

At last we subset the peak area quantification table to features for which we
have also an MS2 spectrum and export this to the *xcms_onlyMS2.txt* file. This
file can be submitted to GNPS [feature-based molecular
networking](https://ccms-ucsd.github.io/GNPSDocumentation/featurebasedmolecularnetworking/):

```{r}
## filter data table to contain only peaks with MSMS DF[ , !(names(DF) %in% drops)]

##!!! NB before mistake? no _maxTic!!!
filteredDataTable <- dataTable[which(
    dataTable$Row.names %in% filteredMs2Spectra_maxTic@elementMetadata$feature_id),]
```

```{r}
head(filteredDataTable)
```

```{r}
write.table(filteredDataTable, "xcms_maxTic_corr0801.txt", sep = "\t", quote = FALSE, row.names = FALSE)



########################################
filteredMs2Spectra_consensus <- combineSpectra(
    filteredMs2Spectra, fcol = "feature_id", method = consensusSpectrum,
    mzd = 0, minProp = 0.8, ppm = 10)

writeMgfData(filteredMs2Spectra_consensus, "ms2spectra_consensus0801corr.mgf")

consensusDataTable <- dataTable[which(dataTable$Row.names %in%
                                      filteredMs2Spectra_consensus@elementMetadata$feature_id),]
write.table(consensusDataTable, "xcms_consensusMS2_0801corr.txt",
            sep = "\t", quote = FALSE, row.names = FALSE)
```
